Finiteness and Cardinality in Homotopy Type Theory

### 2020-01-10 paper.tex
finished

  The other direction is proven in \cite{fruminFiniteSetsHomotopy2018}, so we
  will not describe it here.
\subsection{Topos}
At this point, we see that a ``decidable Kuratowski finite set'' is precisely
equivalent to a cardinal finite set.
From this, we can lift over all of the properties of cardinal finite sets.
In particular, we see that decidable Kuratowski finite sets form a \emph{topos}.
We already have most of the components we need: closure under \(\bot\),
\(\top\), \(\mathbf{Bool}\), \(+\), \(\times\), and \(\rightarrow\).
What remains is the subobject classifier.
\todo[inline]{Topos!}
\section{Infinite Cardinalities} \label{infinite-cardinalities}
In the previous sections we saw different flavours of finiteness which were
really just different flavours of relations to \(\mathbf{Fin}\).
In this section we will see that we can construct a similar classification of
relations to \(\mathbb{N}\), in the form of the countably infinite types.
\subsection{Split Countable Types}
Our first foray into the world of countable types will be a straightforward
analogue to the split enumerable types.
We need change only one element: instead of a support \emph{list}, we instead
have a support \emph{stream}, which is its infinite.
\begin{rm-definition}[Stream]
    \mathbf{Stream}(A) = \llbracket \top \triangleright \text{const} (\mathbb{N}) \rrbracket(A) \simeq \mathbb{N} \rightarrow A
\begin{rm-definition}[Split Countability]
    \mathcal{E}!(A) = \Sigma {(\mathit{xs} : \mathbf{Stream}(A))} , \Pi {(x : A)} , x \in \mathit{xs}
\subsubsection{Closure}
We know that countable infinity is not closed under the exponential (function
arrow), so the only closure we need to prove is \(\Sigma\) to cover all of
what's left.
To do this we have to take a slightly different approach to the functions we
defined before.
Figure~\ref{pairings} illustrates the reason why: previously, we used the
``Cartesian'' product pairing for each support list.
This diverges if the first list is infinite, never exploring anything other than
the first element in the second list.
Instead, we use here the cantor pairing function, which performs a breadth-first
search of the pairings of both lists.
\input{figures/pairing-functions.tex}
\begin{rm-theorem} \label{split-countability-sigma}
  Split countability is closed under \(\Sigma\).
  The main task here is to find a productive (in the coinductive sense) way to
  combine the two support streams.
  A trick which greatly simplifies the algorithm is to first produce a stream of
  lists, which we then concatenate to a single stream.
  This intermediate stream, which we will call \(C\), is generated from the two
  support streams \(\mathit{xs}\) and \(\mathit{ys}\) like so:
    C_i = \left\{ \left( \mathit{xs}_j , \mathit{ys}_{i - j} \right) \mid j \in [0,i] \right\}
  Split countability is closed under non-dependent product and sum.
  Follows from theorem~\ref{split-countability-sigma}.
\subsubsection{Kleene Star}
While we lose some closures with the inclusion of infinite types, we gain some
others.
In particular, we have the Kleene star.
This means, in effect, that we have closure under lists.
  Split countability is closed under Kleene star.
    \mathcal{E}!(A) \rightarrow \mathcal{E}!(\mathbf{List}(A))
  As with the proof of closure under \(\Sigma\), our main task here is to figure
  out a way to arrange the indices such that 
\subsection{Manifest Countability}
As we can quotient out the position information with finite types, so can we with
countable types.
\todo[inline]{fill in rest here}
\section{Practical Uses} \label{practical}
\subsection{Omniscience}
In this section we are interested in restricted forms of the limited principle
of omniscience \cite{myhillErrettBishopFoundations1972}.
\begin{rm-definition}[Limited Principle of Omniscience]
  For any type \(A\) and predicate \(P\) on \(A\), the limited principle of
  omniscience is as follows:
    \left( \Pi {(x : A)} , \mathbf{Dec}(P(x)) \right) \rightarrow \mathbf{Dec} \left( \Sigma {(x : A)} , P(x) \right)
  In other words, for any decidable predicate the existential quantification of
  that predicate is also decidable.
The limited principle of omniscience is non-constructive, but individual types
can themselves satisfy omniscience.
In particular, \emph{finite} types are omniscient.

@@

All of the finiteness predicates we have seen justify exhaustibility.
We will only prove it once, then, for the weakest:
\begin{rm-theorem}
  Kuratowski-finite types are exhaustible.
\end{rm-theorem}
\begin{proof}
  \todo[inline]{Proof}
\end{proof}

Omniscience is stronger than exhaustibility, as we can derive the latter from
the former:
\begin{rm-lemma} \label{omniscient-is-exhaustible}
  Any omniscient type is exhaustible.
\end{rm-lemma}
\begin{proof}
  For decidable propositions, we know the following:
  \begin{equation} \label{lem-conversion}
    \Pi (x : A) , P(x) \leftrightarrow \neg \Sigma (x : A) , \neg P(x) 
  \end{equation}
  To derive exhaustibility from omniscience, then, we run the predicate in its
  negated form, and then subsequently negate the result.
  The resulting decision over \(\neg \Sigma(x : A) , \neg P(x)\) can be
  converted into \(\Pi (x : A) , P(x)\).
\end{proof}
We cannot derive, however, that any exhaustible type is omniscient, as we do not
have the inverse of equation~\ref{lem-conversion}:
\begin{equation} \label{choice-conversion}
  \Sigma (x : A) , P(x) \leftrightarrow \neg \Pi (x : A) , \neg P(x)
\end{equation}
Such an equation would allow us to pick a representative element from any type,
which is therefore non-constructive.
In a sense, equation~\ref{lem-conversion} requires a form of LEM on the
proposition (i.e. requires it to be decidable), whereas
equation~\ref{choice-conversion} requires a form of choice.
Those finiteness predicates which are ordered do in fact give us this form of
choice, so the conversion is valid.
As such, all of the ordered finiteness predicates imply omniscience.
Again, we will prove it only for the weakest.
\begin{rm-theorem}
  Manifest enumerable types are omniscient.
\end{rm-theorem}
\begin{proof}
  \todo[inline]{Proof}
\end{proof}
Finally, we do have a form of omniscience for prop-valued predicates, as they do
not care about the chosen representative.
\begin{rm-theorem}
  Kuratowski finite types are omniscient about prop-valued predicates.
\end{rm-theorem}
\begin{proof}
  \todo[inline]{Proof}
\end{proof}
\subsection{Synthesising Pattern-Matching Proofs}
In particular, they can automate large proofs by analysing every possible case.
In \cite{firsovDependentlyTypedProgramming2015}, the \(\AgdaDatatype{Pauli}\)
group is used as an example.
\ExecuteMetaData[agda/Data/Pauli.tex]{def}
As \(\AgdaDatatype{Pauli}\) has 4 constructors, \(n\)-ary functions on
\(\AgdaDatatype{Pauli}\) may require up to \(4^n\) cases, making even simple
proofs prohibitively verbose.

@@

A full explanation of our implementation is beyond the scope of this work, so we
only present the finished interface, which is used like so:
\ExecuteMetaData[agda/Data/Pauli.tex]{assoc-prf}

\todo[inline]{Synthesise functions}
\todo[inline]{Partial search over infinite types}
\section{Related Work}
\cite{spiwackConstructivelyFinite2010} explored the notion of ``constructive
finite sets'', demonstrating that there is a rich variety of finiteness
predicates in a constructive setting.
Some of those predicates where explored more in
\cite{firsovVariationsNoetherianness2016a},
\cite{parmannInvestigatingStreamlessSets2015}, and
\cite{uustaluFinitenessRationalSequences2017a}.

>>>>

  The other direction is proven in \cite{fruminFiniteSetsHomotopy2018}, so we
  will not describe it here.
\end{proof}
\subsection{Topos}
At this point, we see that a ``decidable Kuratowski finite set'' is precisely
equivalent to a cardinal finite set.
From this, we can lift over all of the properties of cardinal finite sets.
In particular, we see that decidable Kuratowski finite sets form a \emph{topos}.
We already have most of the components we need: closure under \(\bot\),
\(\top\), \(\mathbf{Bool}\), \(+\), \(\times\), and \(\rightarrow\).
What remains is the subobject classifier.
\todo[inline]{Topos!}
\section{Infinite Cardinalities} \label{infinite-cardinalities}
In the previous sections we saw different flavours of finiteness which were
really just different flavours of relations to \(\mathbf{Fin}\).
In this section we will see that we can construct a similar classification of
relations to \(\mathbb{N}\), in the form of the countably infinite types.
\subsection{Split Countable Types}
Our first foray into the world of countable types will be a straightforward
analogue to the split enumerable types.
We need change only one element: instead of a support \emph{list}, we instead
have a support \emph{stream}, which is its infinite.
\begin{rm-definition}[Stream]
  \begin{equation}
    \mathbf{Stream}(A) = \llbracket \top \triangleright \text{const} (\mathbb{N}) \rrbracket(A) \simeq \mathbb{N} \rightarrow A
  \end{equation}
\end{rm-definition}
\begin{rm-definition}[Split Countability]
  \begin{equation}
    \mathcal{E}!(A) = \Sigma {(\mathit{xs} : \mathbf{Stream}(A))} , \Pi {(x : A)} , x \in \mathit{xs}
  \end{equation}
\end{rm-definition}
This type is definitionally equal to it surjection equivalent (\(\mathbb{N}
\twoheadrightarrow ! \; A\)).
\subsubsection{Closure}
We know that countable infinity is not closed under the exponential (function
arrow), so the only closure we need to prove is \(\Sigma\) to cover all of
what's left.
\input{figures/pairing-functions.tex}
\begin{rm-theorem} \label{split-countability-sigma}
  Split countability is closed under \(\Sigma\).
\end{rm-theorem}
\begin{proof}
  The main task here is to figure out a pattern which will pair elements of the
  two support streams without diverging.
  Figure~\ref{cartesian} illustrates why the pairing function we used for lists
  won't work: it needs to exhaust one of the input lists before inspecting
  anything other than the first element of the other.
  Of course, we can't exhaust a stream, so this diverges.

  Instead, we will use the pattern in figure~\ref{cantor}.
  We will produce a stream of lists (where each list is a diagonal), which is
  then concatenated to the final output stream.
  The intermediate stream is sometimes called the discrete convolution of the
  two input streams.
\end{proof}
\subsubsection{Kleene Star}
While we lose some closures with the inclusion of infinite types, we gain some
others.
In particular, we have the Kleene star.
This means that we have closure under lists.
\begin{rm-theorem}
  Split countability is closed under Kleene star.
  \begin{equation}
    \mathcal{E}!(A) \rightarrow \mathcal{E}!(\mathbf{List}(A))
  \end{equation}
\end{rm-theorem}
\begin{proof}
  As with the proof of closure under \(\Sigma\), our main task here is to figure
  out a pattern that reaches any given finite list in finite time.
  Our first step is to split up the problem as we did for \(\Sigma\)-closure: we
  produce an intermediate stream of lists, which we then concatenate to give the
  final result.
  The intermediate stream is constructed like so:
  \begin{equation}
    \mathit{xs}^\star_i = \left\{ \text{map}(\mathit{xs} , x) \mid \mathit{x} \in \mathbf{List}(\mathbb{N}) , \text{sum}(\text{map}(1+, x)) \equiv i \right\}
  \end{equation}
  In other words, the \(i\)th element is all lists such that the sum of the
  successor of the indices in the list is equal to \(i\).
\end{proof}
\subsection{Other Types of Countability}
There are of course other forms of countability available to us: manifest
countability, cardinal, etc.
We have a similar structure of relationships to the finite types, with the
exception that we cannot construct an isomorphism with \(\mathbb{N}\) fiven a
surjection from \(\mathbb{N}\).
\section{Practical Uses} \label{practical}
\subsection{Omniscience}
In this section we are interested in restricted forms of the limited principle
of omniscience \cite{myhillErrettBishopFoundations1972}.
\begin{rm-definition}[Limited Principle of Omniscience]
  For any type \(A\) and predicate \(P\) on \(A\), the limited principle of
  omniscience is as follows:
  \begin{equation}
    \left( \Pi {(x : A)} , \mathbf{Dec}(P(x)) \right) \rightarrow \mathbf{Dec} \left( \Sigma {(x : A)} , P(x) \right)
  \end{equation}
  In other words, for any decidable predicate the existential quantification of
  that predicate is also decidable.
\end{rm-definition}
The limited principle of omniscience is non-constructive, but individual types
can themselves satisfy omniscience.
In particular, \emph{finite} types are omniscient.

@@

All of the finiteness predicates we have seen justify exhaustibility.
We will only prove it once, then, for the weakest:
\begin{rm-theorem}
  Kuratowski-finite types are exhaustible.
\end{rm-theorem}

Omniscience is stronger than exhaustibility, as we can derive the latter from
the former:
\begin{rm-lemma} \label{omniscient-is-exhaustible}
  Any omniscient type is exhaustible.
\end{rm-lemma}
\begin{proof}
  For decidable propositions, we know the following:
  \begin{equation} \label{lem-conversion}
    \Pi (x : A) , P(x) \leftrightarrow \neg \Sigma (x : A) , \neg P(x) 
  \end{equation}
  To derive exhaustibility from omniscience, then, we run the predicate in its
  negated form, and then subsequently negate the result.
  The resulting decision over \(\neg \Sigma(x : A) , \neg P(x)\) can be
  converted into \(\Pi (x : A) , P(x)\).
\end{proof}
We cannot derive, however, that any exhaustible type is omniscient, as we do not
have the inverse of equation~\ref{lem-conversion}:
\begin{equation} \label{choice-conversion}
  \Sigma (x : A) , P(x) \leftrightarrow \neg \Pi (x : A) , \neg P(x)
\end{equation}
Such an equation would allow us to pick a representative element from any type,
which is therefore non-constructive.
In a sense, equation~\ref{lem-conversion} requires a form of LEM on the
proposition (i.e. requires it to be decidable), whereas
equation~\ref{choice-conversion} requires a form of choice.
Those finiteness predicates which are ordered do in fact give us this form of
choice, so the conversion is valid.
As such, all of the ordered finiteness predicates imply omniscience.
Again, we will prove it only for the weakest.
\begin{rm-theorem}
  Manifest enumerable types are omniscient.
\end{rm-theorem}
Finally, we do have a form of omniscience for prop-valued predicates, as they do
not care about the chosen representative.
\begin{rm-theorem}
  Kuratowski finite types are omniscient about prop-valued predicates.
\end{rm-theorem}
\subsection{Synthesising Pattern-Matching Proofs}
In particular, they can automate large proofs by analysing every possible case.
In \cite{firsovDependentlyTypedProgramming2015}, the \(\AgdaDatatype{Pauli}\)
group is used as an example.
\ExecuteMetaData[agda/Data/Pauli.tex]{def}
As \(\AgdaDatatype{Pauli}\) has 4 constructors, \(n\)-ary functions on
\(\AgdaDatatype{Pauli}\) may require up to \(4^n\) cases, making even simple
proofs prohibitively verbose.

@@

A full explanation of our implementation is beyond the scope of this work, so we
only present the finished interface, which is used like so:
\ExecuteMetaData[agda/Data/Pauli.tex]{assoc-prf}
\subsection{Synthesising Functions}
Finally, thanks to extensionality provided to us by HoTT, and the computation
properties provided by CuTT, we can derive decidable equality on functions over
finite types.
We can also use functions in our proof search.
Here, for instance, is a automated procedure which finds the
\(\AgdaFunction{not}\) function on \(\mathbf{Bool}\), given a specification.
\ExecuteMetaData[agda/Data/Pauli.tex]{not-spec}
\section{Related Work}
\cite{spiwackConstructivelyFinite2010} explored the notion of ``constructive
finite sets'', demonstrating that there is a rich variety of finiteness
predicates in a constructive setting.
Some of those predicates where explored more in
\cite{firsovVariationsNoetherianness2016a},
\cite{parmannInvestigatingStreamlessSets2015}, and
\cite{uustaluFinitenessRationalSequences2017a}.
